Revisiting the Relationship between Short-Chain Fatty Acids, the Microbiota, and Colorectal Tumors

### 2018-07-26 submission/manuscript.Rmd
update first paragraph of discussion

The observations from this study do not support the hypothesis that SCFA concentrations are different in individuals with tumors. Whether we directly measured the SCFA concentration or investigated genes associated with their production, no difference could be identified [Figure 1 & 3]. There is an intriguing reason why taxa associated with SCFA production are decreased in CRC but the genes involoved with its' production are not. Mouth-associated microbes such as *Fusobacterium nucleatum* and *Porphyromonas asaccharolytica* have been found to be increased in patients with carcinomas versus individuals without tumors [@Shah2017; @normalization_Sze2017; @meta_analysis_crc_Sze2018]. Both of these bacterial species are known to have strains that can produce SCFAs such as butyrate [@Vital2014]. Thus the reason we may be observing no change in genes involved with SCFA synthesis, as well as no change to SCFAs themselves, is because the production is being supported by more inflammatory microbes associated with CRC. Additionally, our observations that no benefit could be found in using the concentrations to help classify individuals with and without tumors would be consistent with this reason [Figure 2]. However, our observations are in stark contrast to some of the previous literature.

>>>>

The observations from this study do not support the hypothesis that SCFA concentrations are different in individuals with tumors. Whether we directly measured the SCFA concentration or investigated genes associated with their production, no difference could be identified between individuals without tumors and patients with adenomas or carcinomas [Figure 1 & 2]. Although there were differences in the number of significant correlations between SCFA concentration and OTU relative abundance based on whether individuals did not have tumors, had an adenoma, or had a carcinoma, SCFA concentrations did not provide increased model accuracy for tumor classification [Figure 3-4 & S1]. Instead, SCFA concentrations provided similar information to what specific OTUs were already providing to the tumor classification models [Figure 4]. Additionally, when models using OTU relative abundance to classify SCFA concentrations were assessed, the OTUs that classified to *Clostridiales*, *Lachnospiraceae*, and *Ruminococcaceae* were not the same as the OTUs that classified to  these taxa in the tumor models [Figure 4-5]. Collectively, our observations suggest that resident taxa from *Clostridiales*, *Lachnospiraceae*, and *Ruminococcaceae*, that are different between individuals without tumors and patients with adenomas or carinomas, are not that same as those involved with SCFA production.

### 2018-07-26 submission/manuscript.tex
update first paragraph of discussion

The observations from this study do not support the hypothesis that SCFA
concentrations are different in individuals with tumors. Whether we
directly measured the SCFA concentration or investigated genes
associated with their production, no difference could be identified
{[}Figure 1 \& 3{]}. There is an intriguing reason why taxa associated
with SCFA production are decreased in CRC but the genes involoved with
its' production are not. Mouth-associated microbes such as
\emph{Fusobacterium nucleatum} and \emph{Porphyromonas asaccharolytica}
have been found to be increased in patients with carcinomas versus
individuals without tumors (10, 11, 23). Both of these bacterial species
are known to have strains that can produce SCFAs such as butyrate (24).
Thus the reason we may be observing no change in genes involved with
SCFA synthesis, as well as no change to SCFAs themselves, is because the
production is being supported by more inflammatory microbes associated
with CRC. Additionally, our observations that no benefit could be found
in using the concentrations to help classify individuals with and
without tumors would be consistent with this reason {[}Figure 2{]}.
However, our observations are in stark contrast to some of the previous
literature.

Much of the previous research on SCFA benefit to human CRC has been
illustrated in model systems (16). Many SCFAs are produced through the
breakdown of fiber (14) and a recent study in mice found that fiber
supplementation increased SCFA concentrations and decreased tumor burden
(18). Additionally, SCFAs such as butyrate can inhibit tumor growth in
\emph{in vitro} experiments (17). Yet, observations in humans has been
mixed. Previous case/control studies found associations with lower SCFA
concentrations in individuals with carcinoma tumors (19). However,
individual randomized controlled trials and a recent meta-analysis on
fiber supplementation to prevent tumor recurrence has found no benefit
(20, 25). Our results align with what has been reported in
randomized-controlled trials, that SCFAs do not provide general
protection against colorectal tumors. It is possible though that there
are specific instances where SCFAs may be beneficial.

One limitation of current research into the effect of SCFAs in CRC has
been that all tumors are treated as the same type. However, there are
known differences in the types of mutations that occur (26) and treating
all tumors as equal may actually hide any benefit that could be found in
specific subsets of individuals. Similar to the idea of specific
immunotherapy for specific tumors (27), SCFAs may have beneficial
effects for specific types of colorectal tumors. Future research will
need to test if this is a valid hypothesis. Another limitation is that a
fecal sample may not be an ideal type of biospecimen and that the effect
SCFAs have on tumorigenesis is only detected in the colon. However, most
\emph{in vivo} studies as well as human studies have used fecal material
in their analysis (18, 19). Additionally, studies that measure SCFA
changes after fiber supplementation use fecal material to track these
responses (28). Although there are limitations with the current research
on SCFAs and colorectal tumors, our observations along with the
randomized controlled trials on fiber supplementation in tumor
recurrence (20) provide evidence that these specific metabolites may not
be protective. Yet, taxa that are associated with SCFA production are
consistently higher in indivdiuals without colon tumors than patients
with carcinomas (10, 11, 23).

The potential protection against colorectal cancers may not be from
SCFAs even though taxa associated with their production are higher in
individuals without tumors versus patients with carcinomas (10, 11, 23).
Protection could be via a different pathway and by extension other
metabolites that have not been extensively studied. Alternatively,
protection may not occur via a metabolite but instead through niche
exclusion of mouth-associated microbes (e.g. \emph{Fusobacterium},
\emph{Porphyromonas}, \emph{Parvimonas}, \emph{Peptostreptococcus} (6,
12, 13)). The idea of niche exclusion is similar to how the community
protects against \emph{Clostridium difficile} infection (29) with
chronic inflammation replacing the role of antibiotics. Although we did
not find lower concentrations of SCFAs associated with colorectal
tumors, we think that there are many exciting new avenues to explore
because of these results.

@@

\textbf{Study design and sampling.} The overall protocol has been
described in detail previously (30, 31). In brief, this study used fecal
samples obtained at either a single cross-sectional time point (n=490)
or from before (pre-) and after (post-) treatment of a patient's tumor
(n=67). For patients undergoing treatment for their tumor the length of
time between their initial and follow up sample ranged from 188 - 546
days. Our use of treatment has been previously defined as encompassing
removal of a tumor with or without chemotherapy and radiation (30).
Diagnosis of tumor was made by colonoscopic examination and
histopathological review of biopsies obtained (30, 31). The University
of Michigan Institutional Review Board approved the study and informed
consent was obtained from all participants in accordance to the
guidelines set out by the Helsinki Declaration.

\textbf{Measuring specific SCFAs.} Our protocol for the measurement of
acetate, butyrate, and propionate followed a previously published
protocol that used a High-Performance Liquid Chromatography (HPLC)
machine (28). The following changes to this protocol included the use of
frozen fecal samples suspended in 1ml of PBS instead of fecal
suspensions in DNA Genotek OmniGut tubes, and the use of the actual
weight of fecal samples instead of the average weight for SCFA
concentration normalizations. These methodological changes did not
affect the overall median concentrations of these SCFAs between the two
studies (see Table 1 (28) and Figure 1 here).

\textbf{16s rRNA gene sequencing.} The workflow and processing have been
previously described (30, 32, 33). The major differences from these
previous reports include: the use of version 1.39.5 of the mothur
software package and clustering Operational Taxonomic Units (OTUs) at
97\% similarity using the OptClust algorithm (34).

@@

\textbf{Obtaining Operational Protein Families from metagenomes.} A
subset of the cross-sectional group (n=490) containing a total of 85
individuals (normal n=29, adenoma n=28, and carcinoma n=28) was shotgun
sequenced on an Illumina HiSeq using 125 bp paired end reads and a
previously described method (35). Briefly, the sequences were quality
filtered and sequences aligning to the human genome were removed prior
to contig assembly with MEGAHIT (36). Open Reading Frames (ORFs) were
identified using Prodigal (37), counts generated using Diamond (38),
subsequent clustering into Operational Protein Families (OPFs) used
mmseq2 (39), and OPF alignment used the KEGG database (40).

\textbf{Pulling genes involved with SCFA synthesis.} Specific genes
located near the end of the pathways involved in the synthesis of
acetate, butyrate, and propionate were analyzed for any differences
between individuals with normal colons and those with tumors. These
genes were based on pathways from KEGG as well as previous research (40,
41) and a list can be found in the supplemental material {[}Table S1{]}.

@@

\textbf{Statistical analysis workflow.} All analysis was performed using
the statistical language R (42). Generally, a Kruskal-Walis rank sum
test with a Dunn's post-hoc test was used to assess differences between
the groups used. Where appropriate Benjamini-Hochberg was used to
correct for multiple comparisons (43). First, we assessed differences in
SCFA concentrations measured by HPLC between individuals with normal
colons and patients with tumors (adenoma or carcinoma). We then analyzed
whether SCFA concentrations changed in patients with an adenoma or
carcinoma pre- versus post-treatment. Next, we assessed whether OTUs
alone or OTUs and SCFAs were better able to classify individuals with
and without tumor using Random Forest models. Next, the imputed gene
counts of important mediators of SCFA synthesis was tested.
Additionally, the counts generated for OPFs that matched important genes
involved with SCFA creation were analyzed. Finally, models to classify
high or low SCFA concentration based on the median of each SCFA and 16S
rRNA gene sequencing data was created using the Random Forest algorithm.

@@

\hypertarget{ref-normalization_Sze2017}{}
23. \textbf{Sze MA}, \textbf{Baxter NT}, \textbf{Ruffin MT},
\textbf{Rogers MAM}, \textbf{Schloss PD}. 2017. Normalization of the
microbiota in patients after treatment for colonic lesions. Microbiome
\textbf{5}.
doi:\href{https://doi.org/10.1186/s40168-017-0366-3}{10.1186/s40168-017-0366-3}.

\hypertarget{ref-Vital2014}{}
24. \textbf{Vital M}, \textbf{Howe AC}, \textbf{Tiedje JM}. 2014.
Revealing the bacterial butyrate synthesis pathways by analyzing
(meta)genomic data. mBio \textbf{5}:e00889--14--e00889--14.
doi:\href{https://doi.org/10.1128/mbio.00889-14}{10.1128/mbio.00889-14}.

\hypertarget{ref-Schatzkin2000}{}
25. \textbf{Schatzkin A}, \textbf{Lanza E}, \textbf{Corle D},
\textbf{Lance P}, \textbf{Iber F}, \textbf{Caan B}, \textbf{Shike M},
\textbf{Weissfeld J}, \textbf{Burt R}, \textbf{Cooper MR},
\textbf{Kikendall JW}, \textbf{Cahill J}, \textbf{Freedman L},
\textbf{Marshall J}, \textbf{Schoen RE}, \textbf{Slattery M}. 2000. Lack
of effect of a low-fat, high-fiber diet on the recurrence of colorectal
adenomas. New England Journal of Medicine \textbf{342}:1149--1155.
doi:\href{https://doi.org/10.1056/nejm200004203421601}{10.1056/nejm200004203421601}.

\hypertarget{ref-Fearon1990}{}
26. \textbf{Fearon ER}, \textbf{Vogelstein B}. 1990. A genetic model for
colorectal tumorigenesis. Cell \textbf{61}:759--767.
doi:\href{https://doi.org/10.1016/0092-8674(90)90186-i}{10.1016/0092-8674(90)90186-i}.

\hypertarget{ref-Thomas2016}{}
27. \textbf{Thomas X}, \textbf{Heiblig M}. 2016. The development of
agents targeting the BCR-ABL tyrosine kinase as philadelphia
chromosome-positive acute lymphoblastic leukemia treatment. Expert
Opinion on Drug Discovery \textbf{11}:1061--1070.
doi:\href{https://doi.org/10.1080/17460441.2016.1227318}{10.1080/17460441.2016.1227318}.

\hypertarget{ref-scfa_measures_venkataraman2016}{}
28. \textbf{Venkataraman A}, \textbf{Sieber JR}, \textbf{Schmidt AW},
\textbf{Waldron C}, \textbf{Theis KR}, \textbf{Schmidt TM}. 2016.
Variable responses of human microbiomes to dietary supplementation with
resistant starch. Microbiome \textbf{4}.
doi:\href{https://doi.org/10.1186/s40168-016-0178-x}{10.1186/s40168-016-0178-x}.

\hypertarget{ref-Theriot2015}{}
29. \textbf{Theriot CM}, \textbf{Young VB}. 2015. Interactions between
the gastrointestinal microbiome and clostridium difficile. Annual Review
of Microbiology \textbf{69}:445--461.
doi:\href{https://doi.org/10.1146/annurev-micro-091014-104115}{10.1146/annurev-micro-091014-104115}.

\hypertarget{ref-normalization_sze2017}{}
30. \textbf{Sze MA}, \textbf{Baxter NT}, \textbf{Ruffin MT},
\textbf{Rogers MAM}, \textbf{Schloss PD}. 2017. Normalization of the
microbiota in patients after treatment for colonic lesions. Microbiome
\textbf{5}.
doi:\href{https://doi.org/10.1186/s40168-017-0366-3}{10.1186/s40168-017-0366-3}.

\hypertarget{ref-crc_model_baxter2016}{}
31. \textbf{Baxter NT}, \textbf{Ruffin MT}, \textbf{Rogers MAM},
\textbf{Schloss PD}. 2016. Microbiota-based model improves the
sensitivity of fecal immunochemical test for detecting colonic lesions.
Genome Medicine \textbf{8}.
doi:\href{https://doi.org/10.1186/s13073-016-0290-3}{10.1186/s13073-016-0290-3}.

\hypertarget{ref-Schloss2009}{}
32. \textbf{Schloss PD}, \textbf{Westcott SL}, \textbf{Ryabin T},
\textbf{Hall JR}, \textbf{Hartmann M}, \textbf{Hollister EB},
\textbf{Lesniewski RA}, \textbf{Oakley BB}, \textbf{Parks DH},
\textbf{Robinson CJ}, \textbf{Sahl JW}, \textbf{Stres B},
\textbf{Thallinger GG}, \textbf{Horn DJV}, \textbf{Weber CF}. 2009.
Introducing mothur: Open-source, platform-independent,
community-supported software for describing and comparing microbial
communities. Applied and Environmental Microbiology
\textbf{75}:7537--7541.
doi:\href{https://doi.org/10.1128/aem.01541-09}{10.1128/aem.01541-09}.

\hypertarget{ref-Kozich2013}{}
33. \textbf{Kozich JJ}, \textbf{Westcott SL}, \textbf{Baxter NT},
\textbf{Highlander SK}, \textbf{Schloss PD}. 2013. Development of a
dual-index sequencing strategy and curation pipeline for analyzing
amplicon sequence data on the MiSeq illumina sequencing platform.
Applied and Environmental Microbiology \textbf{79}:5112--5120.
doi:\href{https://doi.org/10.1128/aem.01043-13}{10.1128/aem.01043-13}.

\hypertarget{ref-opticlust_westcott2017}{}
34. \textbf{Westcott SL}, \textbf{Schloss PD}. 2017. OptiClust, an
improved method for assigning amplicon-based sequence data to
operational taxonomic units. mSphere \textbf{2}:e00073--17.
doi:\href{https://doi.org/10.1128/mspheredirect.00073-17}{10.1128/mspheredirect.00073-17}.

\hypertarget{ref-Hannigan2017}{}
35. \textbf{Hannigan GD}, \textbf{Duhaime MB}, \textbf{Ruffin MT},
\textbf{Koumpouras CC}, \textbf{Schloss PD}. 2017. Diagnostic potential
\& the interactive dynamics of the colorectal cancer virome.
doi:\href{https://doi.org/10.1101/152868}{10.1101/152868}.

\hypertarget{ref-Li2015}{}
36. \textbf{Li D}, \textbf{Liu C-M}, \textbf{Luo R}, \textbf{Sadakane
K}, \textbf{Lam T-W}. 2015. MEGAHIT: An ultra-fast single-node solution
for large and complex metagenomics assembly via succinct de bruijn
graph. Bioinformatics \textbf{31}:1674--1676.
doi:\href{https://doi.org/10.1093/bioinformatics/btv033}{10.1093/bioinformatics/btv033}.

\hypertarget{ref-Hyatt2010}{}
37. \textbf{Hyatt D}, \textbf{Chen G-L}, \textbf{LoCascio PF},
\textbf{Land ML}, \textbf{Larimer FW}, \textbf{Hauser LJ}. 2010.
Prodigal: Prokaryotic gene recognition and translation initiation site
identification. BMC Bioinformatics \textbf{11}:119.
doi:\href{https://doi.org/10.1186/1471-2105-11-119}{10.1186/1471-2105-11-119}.

\hypertarget{ref-Buchfink2014}{}
38. \textbf{Buchfink B}, \textbf{Xie C}, \textbf{Huson DH}. 2014. Fast
and sensitive protein alignment using DIAMOND. Nature Methods
\textbf{12}:59--60.
doi:\href{https://doi.org/10.1038/nmeth.3176}{10.1038/nmeth.3176}.

\hypertarget{ref-Steinegger2017}{}
39. \textbf{Steinegger M}, \textbf{Söding J}. 2017. MMseqs2 enables
sensitive protein sequence searching for the analysis of massive data
sets. Nature Biotechnology.
doi:\href{https://doi.org/10.1038/nbt.3988}{10.1038/nbt.3988}.

\hypertarget{ref-Kanehisa2015}{}
40. \textbf{Kanehisa M}, \textbf{Sato Y}, \textbf{Kawashima M},
\textbf{Furumichi M}, \textbf{Tanabe M}. 2015. KEGG as a reference
resource for gene and protein annotation. Nucleic Acids Research
\textbf{44}:D457--D462.
doi:\href{https://doi.org/10.1093/nar/gkv1070}{10.1093/nar/gkv1070}.

\hypertarget{ref-scfa_baxter2014}{}
41. \textbf{Baxter NT}, \textbf{Zackular JP}, \textbf{Chen GY},
\textbf{Schloss PD}. 2014. Structure of the gut microbiome following
colonization with human feces determines colonic tumor burden.
Microbiome \textbf{2}:20.
doi:\href{https://doi.org/10.1186/2049-2618-2-20}{10.1186/2049-2618-2-20}.

\hypertarget{ref-r_citation_2017}{}
42. \textbf{R Core Team}. 2017. R: A language and environment for
statistical computing. R Foundation for Statistical Computing, Vienna,
Austria.

\hypertarget{ref-benjamini_controlling_1995}{}
43. \textbf{Benjamini Y}, \textbf{Hochberg Y}. 1995. Controlling the
false discovery rate: A practical and powerful approach to multiple
testing. Journal of the Royal Statistical Society Series B
(Methodological) \textbf{57}:289--300.

>>>>

The observations from this study do not support the hypothesis that SCFA
concentrations are different in individuals with tumors. Whether we
directly measured the SCFA concentration or investigated genes
associated with their production, no difference could be identified
between individuals without tumors and patients with adenomas or
carcinomas {[}Figure 1 \& 2{]}. Although there were differences in the
number of significant correlations between SCFA concentration and OTU
relative abundance based on whether individuals did not have tumors, had
an adenoma, or had a carcinoma, SCFA concentrations did not provide
increased model accuracy for tumor classification {[}Figure 3-4 \&
S1{]}. Instead, SCFA concentrations provided similar information to what
specific OTUs were already providing to the tumor classification models
{[}Figure 4{]}. Additionally, when models using OTU relative abundance
to classify SCFA concentrations were assessed, the OTUs that classified
to \emph{Clostridiales}, \emph{Lachnospiraceae}, and
\emph{Ruminococcaceae} were not the same as the OTUs that classified to
these taxa in the tumor models {[}Figure 4-5{]}. Collectively, our
observations suggest that resident taxa from \emph{Clostridiales},
\emph{Lachnospiraceae}, and \emph{Ruminococcaceae}, that are different
between individuals without tumors and patients with adenomas or
carinomas, are not that same as those involved with SCFA production.

Much of the previous research on SCFA benefit to human CRC has been
illustrated in model systems (16). Many SCFAs are produced through the
breakdown of fiber (14) and a recent study in mice found that fiber
supplementation increased SCFA concentrations and decreased tumor burden
(18). Additionally, SCFAs such as butyrate can inhibit tumor growth in
\emph{in vitro} experiments (17). Yet, observations in humans has been
mixed. Previous case/control studies found associations with lower SCFA
concentrations in individuals with carcinoma tumors (19). However,
individual randomized controlled trials and a recent meta-analysis on
fiber supplementation to prevent tumor recurrence has found no benefit
(20, 23). Our results align with what has been reported in
randomized-controlled trials, that SCFAs do not provide general
protection against colorectal tumors. It is possible though that there
are specific instances where SCFAs may be beneficial.

One limitation of current research into the effect of SCFAs in CRC has
been that all tumors are treated as the same type. However, there are
known differences in the types of mutations that occur (24) and treating
all tumors as equal may actually hide any benefit that could be found in
specific subsets of individuals. Similar to the idea of specific
immunotherapy for specific tumors (25), SCFAs may have beneficial
effects for specific types of colorectal tumors. Future research will
need to test if this is a valid hypothesis. Another limitation is that a
fecal sample may not be an ideal type of biospecimen and that the effect
SCFAs have on tumorigenesis is only detected in the colon. However, most
\emph{in vivo} studies as well as human studies have used fecal material
in their analysis (18, 19). Additionally, studies that measure SCFA
changes after fiber supplementation use fecal material to track these
responses (26). Although there are limitations with the current research
on SCFAs and colorectal tumors, our observations along with the
randomized controlled trials on fiber supplementation in tumor
recurrence (20) provide evidence that these specific metabolites may not
be protective. Yet, taxa that are associated with SCFA production are
consistently higher in indivdiuals without colon tumors than patients
with carcinomas (10, 11, 27).

The potential protection against colorectal cancers may not be from
SCFAs even though taxa associated with their production are higher in
individuals without tumors versus patients with carcinomas (10, 11, 27).
Protection could be via a different pathway and by extension other
metabolites that have not been extensively studied. Alternatively,
protection may not occur via a metabolite but instead through niche
exclusion of mouth-associated microbes (e.g. \emph{Fusobacterium},
\emph{Porphyromonas}, \emph{Parvimonas}, \emph{Peptostreptococcus} (6,
12, 13)). The idea of niche exclusion is similar to how the community
protects against \emph{Clostridium difficile} infection (28) with
chronic inflammation replacing the role of antibiotics. Although we did
not find lower concentrations of SCFAs associated with colorectal
tumors, we think that there are many exciting new avenues to explore
because of these results.

@@

\textbf{Study design and sampling.} The overall protocol has been
described in detail previously (29, 30). In brief, this study used fecal
samples obtained at either a single cross-sectional time point (n=490)
or from before (pre-) and after (post-) treatment of a patient's tumor
(n=67). For patients undergoing treatment for their tumor the length of
time between their initial and follow up sample ranged from 188 - 546
days. Our use of treatment has been previously defined as encompassing
removal of a tumor with or without chemotherapy and radiation (29).
Diagnosis of tumor was made by colonoscopic examination and
histopathological review of biopsies obtained (29, 30). The University
of Michigan Institutional Review Board approved the study and informed
consent was obtained from all participants in accordance to the
guidelines set out by the Helsinki Declaration.

\textbf{Measuring specific SCFAs.} Our protocol for the measurement of
acetate, butyrate, and propionate followed a previously published
protocol that used a High-Performance Liquid Chromatography (HPLC)
machine (26). The following changes to this protocol included the use of
frozen fecal samples suspended in 1ml of PBS instead of fecal
suspensions in DNA Genotek OmniGut tubes, and the use of the actual
weight of fecal samples instead of the average weight for SCFA
concentration normalizations. These methodological changes did not
affect the overall median concentrations of these SCFAs between the two
studies (see Table 1 (26) and Figure 1 here).

\textbf{16s rRNA gene sequencing.} The workflow and processing have been
previously described (29, 31, 32). The major differences from these
previous reports include: the use of version 1.39.5 of the mothur
software package and clustering Operational Taxonomic Units (OTUs) at
97\% similarity using the OptClust algorithm (33).

@@

\textbf{Obtaining Operational Protein Families from metagenomes.} A
subset of the cross-sectional group (n=490) containing a total of 85
individuals (normal n=29, adenoma n=28, and carcinoma n=28) was shotgun
sequenced on an Illumina HiSeq using 125 bp paired end reads and a
previously described method (34). Briefly, the sequences were quality
filtered and sequences aligning to the human genome were removed prior
to contig assembly with MEGAHIT (35). Open Reading Frames (ORFs) were
identified using Prodigal (36), counts generated using Diamond (37),
subsequent clustering into Operational Protein Families (OPFs) used
mmseq2 (38), and OPF alignment used the KEGG database (39).

\textbf{Pulling genes involved with SCFA synthesis.} Specific genes
located near the end of the pathways involved in the synthesis of
acetate, butyrate, and propionate were analyzed for any differences
between individuals with normal colons and those with tumors. These
genes were based on pathways from KEGG as well as previous research (39,
40) and a list can be found in the supplemental material {[}Table S1{]}.

@@

\textbf{Statistical analysis workflow.} All analysis was performed using
the statistical language R (41). Generally, a Kruskal-Walis rank sum
test with a Dunn's post-hoc test was used to assess differences between
the groups used. Where appropriate Benjamini-Hochberg was used to
correct for multiple comparisons (42). First, we assessed differences in
SCFA concentrations measured by HPLC between individuals with normal
colons and patients with tumors (adenoma or carcinoma). We then analyzed
whether SCFA concentrations changed in patients with an adenoma or
carcinoma pre- versus post-treatment. Next, we assessed whether OTUs
alone or OTUs and SCFAs were better able to classify individuals with
and without tumor using Random Forest models. Next, the imputed gene
counts of important mediators of SCFA synthesis was tested.
Additionally, the counts generated for OPFs that matched important genes
involved with SCFA creation were analyzed. Finally, models to classify
high or low SCFA concentration based on the median of each SCFA and 16S
rRNA gene sequencing data was created using the Random Forest algorithm.

@@

\hypertarget{ref-Schatzkin2000}{}
23. \textbf{Schatzkin A}, \textbf{Lanza E}, \textbf{Corle D},
\textbf{Lance P}, \textbf{Iber F}, \textbf{Caan B}, \textbf{Shike M},
\textbf{Weissfeld J}, \textbf{Burt R}, \textbf{Cooper MR},
\textbf{Kikendall JW}, \textbf{Cahill J}, \textbf{Freedman L},
\textbf{Marshall J}, \textbf{Schoen RE}, \textbf{Slattery M}. 2000. Lack
of effect of a low-fat, high-fiber diet on the recurrence of colorectal
adenomas. New England Journal of Medicine \textbf{342}:1149--1155.
doi:\href{https://doi.org/10.1056/nejm200004203421601}{10.1056/nejm200004203421601}.

\hypertarget{ref-Fearon1990}{}
24. \textbf{Fearon ER}, \textbf{Vogelstein B}. 1990. A genetic model for
colorectal tumorigenesis. Cell \textbf{61}:759--767.
doi:\href{https://doi.org/10.1016/0092-8674(90)90186-i}{10.1016/0092-8674(90)90186-i}.

\hypertarget{ref-Thomas2016}{}
25. \textbf{Thomas X}, \textbf{Heiblig M}. 2016. The development of
agents targeting the BCR-ABL tyrosine kinase as philadelphia
chromosome-positive acute lymphoblastic leukemia treatment. Expert
Opinion on Drug Discovery \textbf{11}:1061--1070.
doi:\href{https://doi.org/10.1080/17460441.2016.1227318}{10.1080/17460441.2016.1227318}.

\hypertarget{ref-scfa_measures_venkataraman2016}{}
26. \textbf{Venkataraman A}, \textbf{Sieber JR}, \textbf{Schmidt AW},
\textbf{Waldron C}, \textbf{Theis KR}, \textbf{Schmidt TM}. 2016.
Variable responses of human microbiomes to dietary supplementation with
resistant starch. Microbiome \textbf{4}.
doi:\href{https://doi.org/10.1186/s40168-016-0178-x}{10.1186/s40168-016-0178-x}.

\hypertarget{ref-normalization_Sze2017}{}
27. \textbf{Sze MA}, \textbf{Baxter NT}, \textbf{Ruffin MT},
\textbf{Rogers MAM}, \textbf{Schloss PD}. 2017. Normalization of the
microbiota in patients after treatment for colonic lesions. Microbiome
\textbf{5}.
doi:\href{https://doi.org/10.1186/s40168-017-0366-3}{10.1186/s40168-017-0366-3}.

\hypertarget{ref-Theriot2015}{}
28. \textbf{Theriot CM}, \textbf{Young VB}. 2015. Interactions between
the gastrointestinal microbiome and clostridium difficile. Annual Review
of Microbiology \textbf{69}:445--461.
doi:\href{https://doi.org/10.1146/annurev-micro-091014-104115}{10.1146/annurev-micro-091014-104115}.

\hypertarget{ref-normalization_sze2017}{}
29. \textbf{Sze MA}, \textbf{Baxter NT}, \textbf{Ruffin MT},
\textbf{Rogers MAM}, \textbf{Schloss PD}. 2017. Normalization of the
microbiota in patients after treatment for colonic lesions. Microbiome
\textbf{5}.
doi:\href{https://doi.org/10.1186/s40168-017-0366-3}{10.1186/s40168-017-0366-3}.

\hypertarget{ref-crc_model_baxter2016}{}
30. \textbf{Baxter NT}, \textbf{Ruffin MT}, \textbf{Rogers MAM},
\textbf{Schloss PD}. 2016. Microbiota-based model improves the
sensitivity of fecal immunochemical test for detecting colonic lesions.
Genome Medicine \textbf{8}.
doi:\href{https://doi.org/10.1186/s13073-016-0290-3}{10.1186/s13073-016-0290-3}.

\hypertarget{ref-Schloss2009}{}
31. \textbf{Schloss PD}, \textbf{Westcott SL}, \textbf{Ryabin T},
\textbf{Hall JR}, \textbf{Hartmann M}, \textbf{Hollister EB},
\textbf{Lesniewski RA}, \textbf{Oakley BB}, \textbf{Parks DH},
\textbf{Robinson CJ}, \textbf{Sahl JW}, \textbf{Stres B},
\textbf{Thallinger GG}, \textbf{Horn DJV}, \textbf{Weber CF}. 2009.
Introducing mothur: Open-source, platform-independent,
community-supported software for describing and comparing microbial
communities. Applied and Environmental Microbiology
\textbf{75}:7537--7541.
doi:\href{https://doi.org/10.1128/aem.01541-09}{10.1128/aem.01541-09}.

\hypertarget{ref-Kozich2013}{}
32. \textbf{Kozich JJ}, \textbf{Westcott SL}, \textbf{Baxter NT},
\textbf{Highlander SK}, \textbf{Schloss PD}. 2013. Development of a
dual-index sequencing strategy and curation pipeline for analyzing
amplicon sequence data on the MiSeq illumina sequencing platform.
Applied and Environmental Microbiology \textbf{79}:5112--5120.
doi:\href{https://doi.org/10.1128/aem.01043-13}{10.1128/aem.01043-13}.

\hypertarget{ref-opticlust_westcott2017}{}
33. \textbf{Westcott SL}, \textbf{Schloss PD}. 2017. OptiClust, an
improved method for assigning amplicon-based sequence data to
operational taxonomic units. mSphere \textbf{2}:e00073--17.
doi:\href{https://doi.org/10.1128/mspheredirect.00073-17}{10.1128/mspheredirect.00073-17}.

\hypertarget{ref-Hannigan2017}{}
34. \textbf{Hannigan GD}, \textbf{Duhaime MB}, \textbf{Ruffin MT},
\textbf{Koumpouras CC}, \textbf{Schloss PD}. 2017. Diagnostic potential
\& the interactive dynamics of the colorectal cancer virome.
doi:\href{https://doi.org/10.1101/152868}{10.1101/152868}.

\hypertarget{ref-Li2015}{}
35. \textbf{Li D}, \textbf{Liu C-M}, \textbf{Luo R}, \textbf{Sadakane
K}, \textbf{Lam T-W}. 2015. MEGAHIT: An ultra-fast single-node solution
for large and complex metagenomics assembly via succinct de bruijn
graph. Bioinformatics \textbf{31}:1674--1676.
doi:\href{https://doi.org/10.1093/bioinformatics/btv033}{10.1093/bioinformatics/btv033}.

\hypertarget{ref-Hyatt2010}{}
36. \textbf{Hyatt D}, \textbf{Chen G-L}, \textbf{LoCascio PF},
\textbf{Land ML}, \textbf{Larimer FW}, \textbf{Hauser LJ}. 2010.
Prodigal: Prokaryotic gene recognition and translation initiation site
identification. BMC Bioinformatics \textbf{11}:119.
doi:\href{https://doi.org/10.1186/1471-2105-11-119}{10.1186/1471-2105-11-119}.

\hypertarget{ref-Buchfink2014}{}
37. \textbf{Buchfink B}, \textbf{Xie C}, \textbf{Huson DH}. 2014. Fast
and sensitive protein alignment using DIAMOND. Nature Methods
\textbf{12}:59--60.
doi:\href{https://doi.org/10.1038/nmeth.3176}{10.1038/nmeth.3176}.

\hypertarget{ref-Steinegger2017}{}
38. \textbf{Steinegger M}, \textbf{Söding J}. 2017. MMseqs2 enables
sensitive protein sequence searching for the analysis of massive data
sets. Nature Biotechnology.
doi:\href{https://doi.org/10.1038/nbt.3988}{10.1038/nbt.3988}.

\hypertarget{ref-Kanehisa2015}{}
39. \textbf{Kanehisa M}, \textbf{Sato Y}, \textbf{Kawashima M},
\textbf{Furumichi M}, \textbf{Tanabe M}. 2015. KEGG as a reference
resource for gene and protein annotation. Nucleic Acids Research
\textbf{44}:D457--D462.
doi:\href{https://doi.org/10.1093/nar/gkv1070}{10.1093/nar/gkv1070}.

\hypertarget{ref-scfa_baxter2014}{}
40. \textbf{Baxter NT}, \textbf{Zackular JP}, \textbf{Chen GY},
\textbf{Schloss PD}. 2014. Structure of the gut microbiome following
colonization with human feces determines colonic tumor burden.
Microbiome \textbf{2}:20.
doi:\href{https://doi.org/10.1186/2049-2618-2-20}{10.1186/2049-2618-2-20}.

\hypertarget{ref-r_citation_2017}{}
41. \textbf{R Core Team}. 2017. R: A language and environment for
statistical computing. R Foundation for Statistical Computing, Vienna,
Austria.

\hypertarget{ref-benjamini_controlling_1995}{}
42. \textbf{Benjamini Y}, \textbf{Hochberg Y}. 1995. Controlling the
false discovery rate: A practical and powerful approach to multiple
testing. Journal of the Royal Statistical Society Series B
(Methodological) \textbf{57}:289--300.
